---
title: "ds-2020-final-project"
author: "Ryan Horsey, Scott Engstrom, Pablo Leguizamo, Shashank Koneru"
output: html_document
date: "2026-04-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Description

We are working with a dataset of **42,919 Near-Earth Objects (NEOs)** sourced from NASA's
NeoWs (Near Earth Object Web Service) API. Each row represents one NEO and includes physical
characteristics (estimated diameter, absolute magnitude), orbital properties (eccentricity,
inclination, semi-major axis), close approach data (miss distance, relative velocity), and
a binary classification of whether NASA considers it potentially hazardous.

Our goal is to identify which features best predict hazard classification and build a model
that can accurately classify NEOs as hazardous or non-hazardous.

---

## Data Ingestion

We load the dataset directly from the `.xlsx` file using the `readxl` package. No API calls
are needed at this stage since the data has already been collected and saved locally.

```{r load-data}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)

neo_raw <- read_excel("neo_full_dataset.xlsx")

# Quick look at dimensions and structure
dim(neo_raw)
glimpse(neo_raw)
```

---

## Data Cleaning

### Step 1: Check for Missing Values

Before cleaning, we get a full picture of where data is missing. This tells us which
variables have gaps and helps us decide whether to drop or keep those rows.

```{r missing-values}
colSums(is.na(neo_raw))
```

Two groups of missing values stand out:

- **1,762 rows** are missing both `miss_distance_km` and `relative_velocity_km_s`. These are
  two of our most important predictor variables, so rows missing both cannot contribute
  meaningfully to modeling and will be removed.
- A small number of rows (4–13) are missing diameter and orbital columns. These will also
  be dropped since imputing physical measurements for asteroids is not reliable.

```{r drop-missing}
neo_clean <- neo_raw %>%
  drop_na()

# Confirm no remaining missing values
colSums(is.na(neo_clean))
nrow(neo_clean)
```

### Step 2: Engineer a Mean Diameter Variable

The dataset provides minimum and maximum estimated diameter in kilometers. Rather than
keeping both separately, we engineer a single `diam_avg_km` column as the midpoint
between the two estimates. This gives us one clean numeric predictor representing size.

```{r engineer-diameter}
neo_clean <- neo_clean %>%
  mutate(diam_avg_km = (est_diam_min_km + est_diam_max_km) / 2)
```

### Step 3: Convert the Target Variable to a Factor

The `is_potentially_hazardous_asteroid` column is currently stored as a logical (`TRUE`/`FALSE`).
For modeling and plotting in R, it is better practice to convert this to a **factor** with
meaningful labels. This also makes the variable behave correctly in `ggplot2` legends and
in classification models.

```{r factor-target}
neo_clean <- neo_clean %>%
  mutate(hazardous = factor(
    is_potentially_hazardous_asteroid,
    levels = c(FALSE, TRUE),
    labels = c("Not Hazardous", "Hazardous")
  )) %>%
  select(-is_potentially_hazardous_asteroid)  # remove original logical column
```

### Step 4: Check for Duplicate Records

We verify there are no duplicate NEO IDs in the dataset. Duplicates could inflate model
performance by leaking the same object into both training and test sets.

```{r duplicates}
sum(duplicated(neo_clean$neo_reference_id))
```

No duplicates found. Each row represents a unique NEO.

### Step 5: Final Cleaned Dataset

```{r final-clean}
glimpse(neo_clean)
```

---

## Visualizations

### Class Balance

```{r class-balance}
neo_clean %>%
  count(hazardous) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = hazardous, y = n, fill = hazardous)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.1)), vjust = -0.4) +
  labs(
    title = "Class Distribution of Hazardous Label",
    x = NULL,
    y = "Count"
  ) +
  theme_minimal()
```

### Distributions by Hazard Class

```{r distributions-by-class}
plot_df <- neo_clean %>%
  select(hazardous, diam_avg_km, relative_velocity_km_s, miss_distance_km) %>%
  pivot_longer(
    cols = c(diam_avg_km, relative_velocity_km_s, miss_distance_km),
    names_to = "feature",
    values_to = "value"
  ) %>%
  mutate(
    feature = recode(
      feature,
      diam_avg_km = "Average Diameter (km)",
      relative_velocity_km_s = "Relative Velocity (km/s)",
      miss_distance_km = "Miss Distance (km)"
    )
  )

ggplot(plot_df, aes(x = value, fill = hazardous, color = hazardous)) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  facet_wrap(~ feature, scales = "free_x", ncol = 1) +
  scale_x_log10(labels = scales::label_number()) +
  labs(
    title = "Feature Distributions by Hazard Class (Log X Scale)",
    x = NULL,
    y = "Density",
    fill = "Class",
    color = "Class"
  ) +
  theme_minimal()
```

### Relative Velocity vs. Miss Distance

```{r velocity-vs-miss}
ggplot(
  neo_clean,
  aes(x = relative_velocity_km_s, y = miss_distance_km, color = hazardous)
) +
  geom_point(alpha = 0.45, size = 1.6) +
  scale_y_log10(labels = scales::label_number()) +
  labs(
    title = "Relative Velocity vs. Miss Distance",
    x = "Relative Velocity (km/s)",
    y = "Miss Distance (km, log scale)",
    color = "Class"
  ) +
  theme_minimal()
```

### Correlation Heatmap (Numeric Predictors)

```{r correlation-heatmap}
cor_df <- neo_clean %>%
  select(
    absolute_magnitude_h,
    est_diam_min_km,
    est_diam_max_km,
    diam_avg_km,
    miss_distance_km,
    relative_velocity_km_s,
    orbital_eccentricity,
    orbital_inclination,
    orbital_semi_major_axis
  )

cor_mat <- cor(cor_df, use = "complete.obs")

cor_long <- as.data.frame(as.table(cor_mat)) %>%
  rename(var1 = Var1, var2 = Var2, correlation = Freq)

ggplot(cor_long, aes(x = var1, y = var2, fill = correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Correlation Heatmap of Numeric Predictors",
    x = NULL,
    y = NULL,
    fill = "r"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Variable Importance (Classification Tree)

```{r variable-importance}
library(rpart)

tree_fit <- rpart(
  hazardous ~ diam_avg_km + relative_velocity_km_s + miss_distance_km +
    absolute_magnitude_h + orbital_eccentricity + orbital_inclination +
    orbital_semi_major_axis,
  data = neo_clean,
  method = "class"
)

importance_df <- tibble(
  variable = names(tree_fit$variable.importance),
  importance = as.numeric(tree_fit$variable.importance)
) %>%
  arrange(desc(importance))

ggplot(importance_df, aes(x = reorder(variable, importance), y = importance)) +
  geom_col(fill = "#1B9E77") +
  coord_flip() +
  labs(
    title = "Variable Importance from Classification Tree",
    x = NULL,
    y = "Importance"
  ) +
  theme_minimal()
```
