---
title: "Exploring NASA Potentially Hazardous Asteroid Classifications"
author: "Ryan Horsey, Scott Engstrom, Pablo Leguizamo, Shashank Koneru"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

## Introduction

Near-Earth Objects (NEOs) are asteroids and comets whose orbits bring them near Earth's
orbit. NASA classifies some near-Earth asteroids as Potentially Hazardous Asteroids
(PHAs), meaning their size and orbital path make them worth continued monitoring. The
label sounds alarming, but it does not mean an asteroid is expected to hit Earth. It
means the object is large enough and close enough to deserve attention.

This project explores which physical, orbital, and close-approach measurements separate
asteroids NASA labels as hazardous from those it does not. We use data wrangling and
visualization as the main tools, then use one classification tree as a simple check on
which variables are most useful for reproducing NASA's label.

A key skepticism point is that NASA's PHA definition is partly based on absolute
magnitude, which is related to asteroid size. We therefore compare a full feature set to
a skeptical feature set that removes size and brightness. This gives the report a clear
roadmap: describe the data, explore the main variable patterns, test whether the same
patterns appear in a simple model, and explain the limits of the findings.

## Data

The primary dataset is `neo_full_dataset.xlsx`, a local object-level dataset with one
row per asteroid. We use this file instead of `neo.csv` because `neo.csv` is event-like:
it contains repeated asteroid IDs from multiple close approaches. Using repeated object
rows without grouped splitting could leak the same asteroid into both training and test
sets.

```{r libraries, echo=TRUE}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(rpart)
library(knitr)
```

```{r display-settings}
class_colors <- c(
  "Not Hazardous" = "#4E79A7",
  "Hazardous" = "#E15759"
)

accent_color <- "#59A14F"
neutral_color <- "#BAB0AC"

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "bottom"
    )
)

feature_labels <- c(
  absolute_magnitude_h = "Absolute Magnitude H",
  diam_avg_km = "Average Diameter (km)",
  miss_distance_km = "Miss Distance (km)",
  relative_velocity_km_s = "Relative Velocity (km/s)",
  orbital_eccentricity = "Orbital Eccentricity",
  orbital_inclination = "Orbital Inclination",
  orbital_semi_major_axis = "Orbital Semi-Major Axis"
)

format_percent <- function(x, accuracy = 0.1) {
  scales::percent(x, accuracy = accuracy)
}
```

```{r load-data, echo=TRUE}
data_path <- "neo_full_dataset.xlsx"
if (!file.exists(data_path)) {
  data_path <- "ds-2020-final/neo_full_dataset.xlsx"
}

neo_raw <- read_excel(data_path)

tibble(
  rows = nrow(neo_raw),
  columns = ncol(neo_raw)
) %>%
  kable(caption = "Raw dataset dimensions")
```

### Data Dictionary

```{r data-dictionary}
data_dictionary <- tibble::tribble(
  ~variable, ~source, ~unit, ~role, ~notes,
  "neo_reference_id", "NASA NeoWs/SBDB extraction", "none", "Identifier", "Used only for duplicate checks, not modeling.",
  "name", "NASA NeoWs/SBDB extraction", "none", "Identifier", "Object name; not used for modeling.",
  "absolute_magnitude_h", "NASA NeoWs/SBDB extraction", "magnitude", "Predictor", "Brightness measure related to asteroid size; definition-adjacent.",
  "est_diam_min_km", "NASA NeoWs/SBDB extraction", "kilometers", "Raw input", "Lower estimated diameter bound; replaced by average diameter.",
  "est_diam_max_km", "NASA NeoWs/SBDB extraction", "kilometers", "Raw input", "Upper estimated diameter bound; replaced by average diameter.",
  "diam_avg_km", "Engineered", "kilometers", "Predictor", "Mean of minimum and maximum estimated diameter.",
  "miss_distance_km", "NASA NeoWs/SBDB extraction", "kilometers", "Predictor", "Close-approach miss distance; related to closeness but not the same as MOID.",
  "relative_velocity_km_s", "NASA NeoWs/SBDB extraction", "km/s", "Predictor", "Relative velocity during close approach.",
  "orbital_eccentricity", "NASA SBDB extraction", "none", "Predictor", "Orbital shape.",
  "orbital_inclination", "NASA SBDB extraction", "degrees", "Predictor", "Tilt of orbit relative to reference plane.",
  "orbital_semi_major_axis", "NASA SBDB extraction", "AU", "Predictor", "Average orbital distance scale.",
  "is_potentially_hazardous_asteroid", "NASA NeoWs/SBDB extraction", "logical", "Target", "Converted to the factor variable hazardous."
)

data_dictionary %>%
  kable(caption = "Data dictionary for variables used in the report")
```

### Missingness and Duplicates

```{r missing-values}
missing_summary <- tibble(
  variable = names(neo_raw),
  missing_n = colSums(is.na(neo_raw)),
  missing_pct = missing_n / nrow(neo_raw)
)

missing_summary %>%
  mutate(missing_pct = format_percent(missing_pct)) %>%
  arrange(desc(missing_n)) %>%
  kable(caption = "Missing values by variable")
```

```{r duplicate-check}
raw_duplicate_ids <- sum(duplicated(neo_raw$neo_reference_id))
raw_unique_ids <- n_distinct(neo_raw$neo_reference_id)

tibble(
  rows = nrow(neo_raw),
  unique_neo_reference_ids = raw_unique_ids,
  duplicate_neo_reference_ids = raw_duplicate_ids
) %>%
  kable(caption = "Duplicate check for the object-level dataset")
```

```{r optional-csv-context}
csv_path <- "../neo.csv"
if (!file.exists(csv_path)) {
  csv_path <- "neo.csv"
}

csv_context <- if (file.exists(csv_path)) {
  neo_csv_context <- read.csv(csv_path)
  tibble(
    dataset = "neo.csv",
    rows = nrow(neo_csv_context),
    unique_ids = n_distinct(neo_csv_context$id),
    interpretation = "Event-like table with repeated asteroid IDs"
  )
} else {
  tibble(
    dataset = "neo.csv",
    rows = NA_integer_,
    unique_ids = NA_integer_,
    interpretation = "Optional context file not found during knit"
  )
}

csv_context %>%
  kable(caption = "Why the event-like CSV was not used as the main modeling table")
```

### Cleaning and Feature Engineering

```{r clean-data, echo=TRUE}
model_vars_full <- c(
  "absolute_magnitude_h",
  "diam_avg_km",
  "miss_distance_km",
  "relative_velocity_km_s",
  "orbital_eccentricity",
  "orbital_inclination",
  "orbital_semi_major_axis"
)

model_vars_skeptical <- c(
  "miss_distance_km",
  "relative_velocity_km_s",
  "orbital_eccentricity",
  "orbital_inclination",
  "orbital_semi_major_axis"
)

neo_clean <- neo_raw %>%
  mutate(
    diam_avg_km = (est_diam_min_km + est_diam_max_km) / 2,
    hazardous = factor(
      is_potentially_hazardous_asteroid,
      levels = c(FALSE, TRUE),
      labels = c("Not Hazardous", "Hazardous")
    )
  ) %>%
  select(
    neo_reference_id,
    hazardous,
    all_of(model_vars_full)
  ) %>%
  drop_na(all_of(model_vars_full), hazardous)

tibble(
  raw_rows = nrow(neo_raw),
  cleaned_rows = nrow(neo_clean),
  rows_removed = nrow(neo_raw) - nrow(neo_clean),
  duplicate_ids_after_cleaning = sum(duplicated(neo_clean$neo_reference_id))
) %>%
  mutate(
    rows_removed_pct = format_percent(rows_removed / raw_rows)
  ) %>%
  select(raw_rows, cleaned_rows, rows_removed, rows_removed_pct, duplicate_ids_after_cleaning) %>%
  kable(caption = "Rows retained after cleaning and feature engineering")
```

## Exploratory Analysis

### Class Balance

```{r class-balance}
class_balance <- neo_clean %>%
  count(hazardous) %>%
  mutate(prop = n / sum(n))

class_balance %>%
  mutate(prop = format_percent(prop)) %>%
  kable(caption = "Class balance in the cleaned asteroid dataset")

ggplot(class_balance, aes(x = hazardous, y = n, fill = hazardous)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(
    aes(label = scales::percent(prop, accuracy = 0.1)),
    vjust = -0.4
  ) +
  labs(
    title = "Class Balance of NASA Hazardous Label",
    x = NULL,
    y = "Asteroid Count"
  ) +
  scale_fill_manual(values = class_colors)
```

The hazardous class is much smaller than the non-hazardous class, so accuracy alone is
not a reliable evaluation metric. A model can look accurate while missing many hazardous
objects.

### Feature Distributions by Hazard Class

```{r distributions-by-class}
distribution_df <- neo_clean %>%
  select(
    hazardous,
    absolute_magnitude_h,
    diam_avg_km,
    miss_distance_km,
    relative_velocity_km_s,
    orbital_eccentricity,
    orbital_inclination,
    orbital_semi_major_axis
  ) %>%
  pivot_longer(
    cols = -hazardous,
    names_to = "feature",
    values_to = "value"
  ) %>%
  mutate(
    feature = recode(feature, !!!feature_labels)
  )

ggplot(distribution_df, aes(x = value, fill = hazardous, color = hazardous)) +
  geom_density(alpha = 0.25, linewidth = 0.7) +
  facet_wrap(~ feature, scales = "free", ncol = 2) +
  labs(
    title = "Predictor Distributions by Hazard Class",
    x = NULL,
    y = "Density",
    fill = "Class",
    color = "Class"
  ) +
  scale_fill_manual(values = class_colors) +
  scale_color_manual(values = class_colors)
```

The clearest separation appears in size-related variables: hazardous asteroids tend to
have lower absolute magnitude values and larger estimated diameters. The orbital and
close-approach variables show more overlap, which suggests that NASA's label is not
explained by one simple measurement in this dataset.

### Relative Velocity vs. Miss Distance

```{r velocity-vs-miss}
ggplot(
  neo_clean,
  aes(x = relative_velocity_km_s, y = miss_distance_km, color = hazardous)
) +
  geom_point(alpha = 0.45, size = 1.5) +
  scale_y_log10(labels = scales::label_number()) +
  labs(
    title = "Relative Velocity vs. Close-Approach Miss Distance",
    x = "Relative Velocity (km/s)",
    y = "Miss Distance (km, log scale)",
    color = "Class"
  ) +
  scale_color_manual(values = class_colors)
```

This scatter plot shows why the problem is not as simple as "closer asteroids are always
hazardous." Hazardous and non-hazardous asteroids overlap across much of the velocity and
miss-distance range.

### Correlation Heatmap

```{r correlation-heatmap}
cor_df <- neo_clean %>%
  select(all_of(model_vars_full))

cor_mat <- cor(cor_df, use = "complete.obs")

cor_long <- as.data.frame(as.table(cor_mat)) %>%
  rename(var1 = Var1, var2 = Var2, correlation = Freq) %>%
  mutate(
    var1 = recode(as.character(var1), !!!feature_labels),
    var2 = recode(as.character(var2), !!!feature_labels)
  )

ggplot(cor_long, aes(x = var1, y = var2, fill = correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = class_colors[["Not Hazardous"]],
    mid = "white",
    high = class_colors[["Hazardous"]],
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Correlation Heatmap of Numeric Predictors",
    x = NULL,
    y = NULL,
    fill = "r"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Average diameter and absolute magnitude are strongly related, which matches the science:
brighter asteroids are usually larger. This correlation is also part of why we avoid
treating size and brightness as independent evidence in the skeptical model below.

### Analysis Decisions and Trials

During earlier drafts, we tried a more model-heavy version of the analysis with logistic
regression, classification trees, ROC curves, and precision-recall curves. Those tools
did not change the main finding: size and brightness carry much of the signal, while the
remaining orbital and close-approach variables are weaker by themselves. We therefore
kept the final report focused on visual exploration and one interpretable classification
tree. We also chose the object-level Excel file over the event-like CSV because repeated
asteroid IDs in the CSV could make model evaluation look better than it really is.

## Modeling

We use a classification tree because it is easier to explain than a more mathematical
model: the tree sorts asteroids by splitting on variables that best separate hazardous
from non-hazardous labels. The tree is fit twice:

- **Full feature set:** orbital, close-approach, size, and brightness predictors.
- **Skeptical feature set:** orbital and close-approach predictors only.

The skeptical feature set tests whether the model can still reproduce NASA's label when
we remove the variables most closely connected to the official PHA definition.

```{r train-test-split, echo=TRUE}
set.seed(2026)

neo_model <- neo_clean %>%
  mutate(row_id = row_number())

train_ids <- neo_model %>%
  group_by(hazardous) %>%
  slice_sample(prop = 0.80) %>%
  ungroup() %>%
  pull(row_id)

train_data <- neo_model %>%
  filter(row_id %in% train_ids) %>%
  select(-row_id, -neo_reference_id)

test_data <- neo_model %>%
  filter(!row_id %in% train_ids) %>%
  select(-row_id, -neo_reference_id)

tibble(
  split = c("Train", "Test"),
  rows = c(nrow(train_data), nrow(test_data)),
  hazardous_rate = c(
    mean(train_data$hazardous == "Hazardous"),
    mean(test_data$hazardous == "Hazardous")
  )
) %>%
  mutate(hazardous_rate = format_percent(hazardous_rate)) %>%
  kable(caption = "Stratified train/test split")
```

### Evaluation Functions

We use four confusion-matrix metrics. Accuracy is the percent of all predictions that
are correct. Recall is the percent of truly hazardous asteroids the model catches.
Specificity is the percent of non-hazardous asteroids the model correctly leaves alone.
Precision is the percent of predicted hazardous asteroids that are actually hazardous.

```{r metric-functions, echo=TRUE}
safe_divide <- function(numerator, denominator) {
  ifelse(denominator == 0, NA_real_, numerator / denominator)
}

confusion_metrics <- function(actual, predicted, positive = "Hazardous") {
  actual <- as.character(actual)
  predicted <- as.character(predicted)
  negative <- setdiff(unique(actual), positive)[1]

  tp <- sum(actual == positive & predicted == positive)
  tn <- sum(actual == negative & predicted == negative)
  fp <- sum(actual == negative & predicted == positive)
  fn <- sum(actual == positive & predicted == negative)

  tibble(
    tp = tp,
    tn = tn,
    fp = fp,
    fn = fn,
    accuracy = safe_divide(tp + tn, tp + tn + fp + fn),
    recall = safe_divide(tp, tp + fn),
    specificity = safe_divide(tn, tn + fp),
    precision = safe_divide(tp, tp + fp)
  )
}
```

### Fit Classification Trees

```{r fit-models, echo=TRUE}
full_formula <- as.formula(
  paste("hazardous ~", paste(model_vars_full, collapse = " + "))
)

skeptical_formula <- as.formula(
  paste("hazardous ~", paste(model_vars_skeptical, collapse = " + "))
)

tree_full <- rpart(
  full_formula,
  data = train_data,
  method = "class",
  control = rpart.control(cp = 0.001, minbucket = 25)
)

tree_skeptical <- rpart(
  skeptical_formula,
  data = train_data,
  method = "class",
  control = rpart.control(cp = 0.001, minbucket = 25)
)

tree_full_predicted <- predict(tree_full, newdata = test_data, type = "class")
tree_skeptical_predicted <- predict(tree_skeptical, newdata = test_data, type = "class")
```

### Slide-Ready Classification Tree

For the presentation, we use a shallower version of the full-feature tree. This is not a
replacement for the full model above; it is a readable visual that shows the first few
splits the tree uses to separate hazardous from non-hazardous asteroids.

```{r slide-classification-tree, fig.width=10, fig.height=5.6}
tree_slide <- rpart(
  full_formula,
  data = train_data,
  method = "class",
  control = rpart.control(
    cp = 0.01,
    maxdepth = 3,
    minbucket = 50
  )
)

plot_slide_tree <- function(tree) {
  if (requireNamespace("rpart.plot", quietly = TRUE)) {
    rpart.plot::rpart.plot(
      tree,
      type = 2,
      extra = 104,
      fallen.leaves = TRUE,
      box.palette = c(
        class_colors[["Not Hazardous"]],
        class_colors[["Hazardous"]]
      ),
      branch.lty = 3,
      shadow.col = "gray",
      cex = 0.72,
      main = "Simple Classification Tree"
    )
  } else {
    plot(tree, uniform = TRUE, margin = 0.08, main = "Simple Classification Tree")
    text(tree, use.n = TRUE, all = TRUE, cex = 0.65)
  }
}

plot_slide_tree(tree_slide)

png("../classification_tree_slide.png", width = 1800, height = 1000, res = 200)
plot_slide_tree(tree_slide)
dev.off()
```

This visual is meant for a quick modeling slide: full feature set, 80/20 train-test
split, and an interpretable tree that predicts NASA's hazardous label.

## Results

### Tree Comparison

```{r model-comparison}
model_comparison <- bind_rows(
  confusion_metrics(test_data$hazardous, tree_full_predicted) %>%
    mutate(feature_set = "Full"),
  confusion_metrics(test_data$hazardous, tree_skeptical_predicted) %>%
    mutate(feature_set = "Skeptical")
) %>%
  select(
    feature_set,
    accuracy,
    recall,
    specificity,
    precision,
    tp,
    tn,
    fp,
    fn
  ) %>%
  arrange(desc(recall), desc(precision))

model_comparison %>%
  mutate(
    across(c(accuracy, recall, specificity, precision), ~ format_percent(.x)),
    note = if_else(feature_set == first(feature_set), "**Highest recall**", "")
  ) %>%
  select(
    `Feature Set` = feature_set,
    Accuracy = accuracy,
    Recall = recall,
    Specificity = specificity,
    Precision = precision,
    `True Positives` = tp,
    `True Negatives` = tn,
    `False Positives` = fp,
    `False Negatives` = fn,
    Note = note
  ) %>%
  kable(caption = "Classification tree performance on the test data", escape = FALSE)
```

The full tree performs better at identifying hazardous asteroids. The skeptical tree has
similar overall accuracy, but it catches none of the hazardous objects in the test set.
This is an important result: without size and brightness, the tree mostly learns to
predict the majority "Not Hazardous" class.

### Confusion Matrix Heatmap

The heatmap below uses the full classification tree, which has the best recall for
hazardous asteroids.

```{r confusion-heatmap}
best_model <- model_comparison %>%
  slice_max(recall, n = 1, with_ties = FALSE)

best_predicted <- if (best_model$feature_set == "Full") {
  tree_full_predicted
} else {
  tree_skeptical_predicted
}

confusion_df <- tibble(
  actual = test_data$hazardous,
  predicted = best_predicted
) %>%
  count(actual, predicted) %>%
  complete(
    actual = levels(test_data$hazardous),
    predicted = levels(best_predicted),
    fill = list(n = 0)
  )

ggplot(confusion_df, aes(x = predicted, y = actual, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), color = "white", fontface = "bold") +
  scale_fill_gradient(low = "#C6DBEF", high = class_colors[["Not Hazardous"]]) +
  labs(
    title = paste("Confusion Matrix: Classification Tree -", best_model$feature_set),
    x = "Predicted Class",
    y = "Actual Class",
    fill = "Count"
  )
```

The full tree correctly identifies some hazardous asteroids, but it still misses many of
them. This keeps the result honest: the model helps show which variables matter, but it
would not be acceptable as a real-world asteroid warning system.

### Classification Tree Variable Importance

```{r variable-importance}
importance_df <- tibble(
  variable = names(tree_full$variable.importance),
  importance = as.numeric(tree_full$variable.importance)
) %>%
  mutate(variable = recode(variable, !!!feature_labels)) %>%
  arrange(desc(importance))

importance_df %>%
  mutate(importance = round(importance, 1)) %>%
  kable(caption = "Most important variables in the full classification tree")

ggplot(importance_df, aes(x = reorder(variable, importance), y = importance)) +
  geom_col(fill = accent_color) +
  coord_flip() +
  labs(
    title = "Variable Importance from Full Classification Tree",
    x = NULL,
    y = "Importance"
  )
```

The model's most important variables reinforce the visual findings: size and brightness
are central to the hazardous label, while the other orbital variables provide additional
but weaker separation.

## Skepticism and Limitations

The strongest limitation is that the hazardous label is not an independent scientific
outcome like a future impact. It is NASA's classification, and that classification is
partly based on variables related to asteroid size and brightness. NASA/JPL CNEOS defines
PHAs using Earth MOID of 0.05 AU or less and absolute magnitude H of 22.0 or brighter
([CNEOS NEO Groups](https://cneos.jpl.nasa.gov/about/neo_groups.html)). Absolute
magnitude is therefore definition-adjacent, and average diameter is closely related to
the same idea.

The dataset also does not include Earth MOID, the official orbital-distance criterion
used in the PHA definition. We use `miss_distance_km`, but close-approach miss distance is
not the same as MOID. This means the model approximates NASA's classification using the
available object-level variables rather than exactly applying NASA's rule.

Other limitations:

- The hazardous class is imbalanced, so accuracy can be misleading.
- Rows with missing modeling fields were removed rather than imputed.
- The close-approach variables appear to summarize one approach field, not the full future
  distribution of possible close approaches.
- The models predict NASA's label; they do not estimate causal asteroid danger or actual
  Earth impact probability.

## Conclusion

The strongest pattern in the data is that NASA's hazardous label is closely tied to
asteroid size and brightness. Hazardous asteroids are rare in the cleaned dataset, making
up about 6.6% of observations, and the full classification tree catches more hazardous
objects than the skeptical tree. When size and brightness are removed, the model mostly
defaults to predicting the majority non-hazardous class.

The main takeaway is therefore not that we built a danger-prediction system. Instead, we
showed that NASA's PHA label can be partly reproduced from available asteroid
measurements, while also showing why that result must be interpreted carefully. The model
is strongest when it includes variables that are close to NASA's own definition.

Future work should add Earth MOID, the official orbital-distance part of the PHA
definition, and should compare multiple close approaches for the same asteroid rather
than relying on one summarized approach field. It would also be useful to separate
asteroids by size group or orbit type to see whether the same patterns hold within more
specific asteroid categories.
### Orbital Characteristics by Hazard Class

```{r orbital-boxplots}
neo_clean %>%
  select(hazardous, orbital_eccentricity, orbital_inclination, orbital_semi_major_axis) %>%
  # log10 transform semi-major axis before pivoting — raw values extend to 300+ AU
  # which collapses the box to nothing; log scale reveals the actual distribution
  mutate(orbital_semi_major_axis = log10(orbital_semi_major_axis)) %>%
  pivot_longer(
    cols = c(orbital_eccentricity, orbital_inclination, orbital_semi_major_axis),
    names_to = "feature",
    values_to = "value"
  ) %>%
  mutate(
    feature = recode(
      feature,
      orbital_eccentricity    = "Orbital Eccentricity",
      orbital_inclination     = "Orbital Inclination (deg)",
      orbital_semi_major_axis = "Semi-Major Axis (log10 AU)"
    )
  ) %>%
  ggplot(aes(x = hazardous, y = value, fill = hazardous)) +
  geom_boxplot(outlier.alpha = 0.15, outlier.size = 0.8) +
  facet_wrap(~ feature, scales = "free_y", ncol = 3) +
  labs(
    title = "Orbital Characteristics by Hazard Classification",
    x = NULL,
    y = NULL,
    fill = "Class"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Absolute Magnitude by Hazard Class

```{r magnitude-density}
ggplot(neo_clean, aes(x = absolute_magnitude_h, fill = hazardous, color = hazardous)) +
  geom_density(alpha = 0.3, linewidth = 0.9) +
  labs(
    title = "Distribution of Absolute Magnitude by Hazard Classification",
    x = "Absolute Magnitude (H)",
    y = "Density",
    fill = "Class",
    color = "Class"
  ) +
  theme_minimal()
```

```{r magnitude-boxplot}
ggplot(neo_clean, aes(x = hazardous, y = absolute_magnitude_h, fill = hazardous)) +
  geom_boxplot(outlier.alpha = 0.2, outlier.size = 0.8, show.legend = FALSE) +
  labs(
    title = "Absolute Magnitude by Hazard Classification",
    x = NULL,
    y = "Absolute Magnitude (H)"
  ) +
  theme_minimal()
```

### Pairwise Scatterplots of Top Predictors

```{r pairwise-scatter}
var_pairs <- list(
  list(x = "diam_avg_km",          y = "absolute_magnitude_h"),
  list(x = "diam_avg_km",          y = "orbital_eccentricity"),
  list(x = "diam_avg_km",          y = "orbital_inclination"),
  list(x = "absolute_magnitude_h", y = "orbital_eccentricity"),
  list(x = "absolute_magnitude_h", y = "orbital_inclination"),
  list(x = "orbital_eccentricity", y = "orbital_inclination")
)

pair_df <- bind_rows(lapply(var_pairs, function(p) {
  neo_clean %>%
    transmute(
      hazardous,
      x_val   = .data[[p$x]],
      y_val   = .data[[p$y]],
      panel   = paste(p$x, "vs", p$y)
    )
}))

ggplot(pair_df, aes(x = x_val, y = y_val, color = hazardous)) +
  geom_point(alpha = 0.2, size = 0.6) +
  facet_wrap(~ panel, scales = "free", ncol = 3) +
  labs(
    title = "Pairwise Scatterplots of Top Four Predictors",
    x = NULL,
    y = NULL,
    color = "Class"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text      = element_text(size = 7),
    axis.text       = element_text(size = 6)
  )
```
