---
title: "Predicting Hazardous Near-Earth Objects"
author: "Ryan Horsey, Scott Engstrom, Pablo Leguizamo, Shashank Koneru"
output: github_document
---

*2026-04-24*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

Every day, NASA's Jet Propulsion Laboratory tracks thousands of asteroids whose orbits
bring them close to Earth. A subset of these Near-Earth Objects (NEOs) are officially
classified as **Potentially Hazardous Asteroids (PHAs)**, objects large enough and
close enough to Earth's orbit that a future impact cannot be completely ruled out.
Understanding what distinguishes a hazardous NEO from a non-hazardous one is both
scientifically interesting and practically important for planetary defense planning.

This report investigates a dataset of 42,919 NEOs sourced from NASA's NeoWs (Near Earth
Object Web Service) API. We explore which physical and orbital characteristics best
separate hazardous from non-hazardous objects, and build a classification tree model to
predict hazard status from those features.

The report is organized as follows. We begin with a description and cleaning of the raw
data, followed by exploratory visualizations covering size, brightness, orbital geometry,
and close-approach dynamics. We then fit a classification tree model and evaluate its
performance on a held-out test set. We close with conclusions and directions for future work.

---

## Data Description

We are working with a dataset of **42,919 Near-Earth Objects (NEOs)** sourced from NASA's
NeoWs (Near Earth Object Web Service) API. Each row represents one NEO and includes physical
characteristics (estimated diameter, absolute magnitude), orbital properties (eccentricity,
inclination, semi-major axis), close approach data (miss distance, relative velocity), and
a binary classification of whether NASA considers it potentially hazardous.

Our goal is to identify which features best predict hazard classification and build a model
that can accurately classify NEOs as hazardous or non-hazardous.

The key variables in the dataset are described below:

| Variable | Type | Description |
|---|---|---|
| `neo_reference_id` | ID | Unique NASA identifier for each NEO |
| `absolute_magnitude_h` | Numeric | Intrinsic brightness (lower = brighter/larger) |
| `est_diam_min_km` / `est_diam_max_km` | Numeric | Estimated diameter range in kilometers |
| `relative_velocity_km_s` | Numeric | Speed relative to Earth at closest approach |
| `miss_distance_km` | Numeric | Distance from Earth at closest approach |
| `orbital_eccentricity` | Numeric | How elliptical the orbit is (0 = circular) |
| `orbital_inclination` | Numeric | Tilt of orbit relative to Earth's orbital plane (degrees) |
| `orbital_semi_major_axis` | Numeric | Average orbital distance from the Sun (AU) |
| `is_potentially_hazardous_asteroid` | Logical | NASA's official hazard classification (our target variable) |

---

## Data Ingestion

The raw data was collected directly from NASA's NeoWs API by a team member prior to
this analysis. Rather than using the Feed endpoint (which returns NEOs by date range and
requires looping over many weekly windows), we used the **Browse endpoint**, which
paginates through NASA's full NEO database in one systematic pass:

```
GET https://api.nasa.gov/neo/rest/v1/neo/browse?page={n}&size=20&api_key=KEY
```

The API returned a total of 2,146 pages at 20 records per page. Because the response is
deeply nested JSON, each page was parsed manually using `purrr::map_*` functions to
extract physical characteristics, close approach data, and orbital fields into a flat
tibble. After all pages were collected and row-bound, the result was exported to
`neo_full_dataset.xlsx` using `openxlsx`. The full extraction script is available in
`DS2020_Final_Data_Extraction.Rmd` in this repository.

The `.xlsx` file is loaded here for analysis using `readxl`. No API calls are made
during knitting, so the report is fully reproducible from the saved dataset.

```{r load-data}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(rpart)
library(knitr)

class_colors <- c(
  "Not Hazardous" = "#4E79A7",
  "Hazardous" = "#E15759"
)

accent_color <- "#59A14F"

feature_labels <- c(
  absolute_magnitude_h = "Absolute Magnitude H",
  diam_avg_km = "Average Diameter (km)",
  miss_distance_km = "Miss Distance (km)",
  relative_velocity_km_s = "Relative Velocity (km/s)",
  orbital_eccentricity = "Orbital Eccentricity",
  orbital_inclination = "Orbital Inclination",
  orbital_semi_major_axis = "Orbital Semi-Major Axis"
)

format_percent <- function(x, accuracy = 0.1) {
  scales::percent(x, accuracy = accuracy)
}

neo_raw <- read_excel("neo_full_dataset.xlsx")

# Quick look at dimensions and structure
dim(neo_raw)
glimpse(neo_raw)
```

---

## Data Cleaning

### Step 1: Check for Missing Values

Before cleaning, we get a full picture of where data is missing. This tells us which
variables have gaps and helps us decide whether to drop or keep those rows.

```{r missing-values}
colSums(is.na(neo_raw))
```

Two groups of missing values stand out:

- **1,762 rows** are missing both `miss_distance_km` and `relative_velocity_km_s`. These are
  two of our most important predictor variables, so rows missing both cannot contribute
  meaningfully to modeling and will be removed.
- A small number of rows (4–13) are missing diameter and orbital columns. These will also
  be dropped since imputing physical measurements for asteroids is not reliable.
- The `name` column has a few missing values, but it is an identifier field rather than
  an analysis variable, so it is not used to decide which rows to keep.

```{r drop-missing}
required_analysis_cols <- c(
  "neo_reference_id",
  "absolute_magnitude_h",
  "est_diam_min_km",
  "est_diam_max_km",
  "miss_distance_km",
  "relative_velocity_km_s",
  "orbital_eccentricity",
  "orbital_inclination",
  "orbital_semi_major_axis",
  "is_potentially_hazardous_asteroid"
)

neo_clean <- neo_raw %>%
  drop_na(all_of(required_analysis_cols))

# Confirm no remaining missing values in analysis columns
colSums(is.na(select(neo_clean, all_of(required_analysis_cols))))
nrow(neo_clean)

tibble(
  dataset = c("Raw data", "Cleaned data"),
  rows = c(nrow(neo_raw), nrow(neo_clean)),
  hazardous_rate = c(
    mean(neo_raw$is_potentially_hazardous_asteroid),
    mean(neo_clean$is_potentially_hazardous_asteroid)
  )
) %>%
  mutate(hazardous_rate = format_percent(hazardous_rate)) %>%
  kable(caption = "Class balance before and after missing-value filtering")
```

### Step 2: Engineer a Mean Diameter Variable

The dataset provides minimum and maximum estimated diameter in kilometers. Rather than
keeping both separately, we engineer a single `diam_avg_km` column as the midpoint
between the two estimates. This gives us one clean numeric predictor representing size.

```{r engineer-diameter}
neo_clean <- neo_clean %>%
  mutate(diam_avg_km = (est_diam_min_km + est_diam_max_km) / 2)
```

### Step 3: Convert the Target Variable to a Factor

The `is_potentially_hazardous_asteroid` column is currently stored as a logical (`TRUE`/`FALSE`).
For modeling and plotting in R, it is better practice to convert this to a **factor** with
meaningful labels. This also makes the variable behave correctly in `ggplot2` legends and
in classification models.

```{r factor-target}
neo_clean <- neo_clean %>%
  mutate(hazardous = factor(
    is_potentially_hazardous_asteroid,
    levels = c(FALSE, TRUE),
    labels = c("Not Hazardous", "Hazardous")
  )) %>%
  select(-is_potentially_hazardous_asteroid)  # remove original logical column
```

### Step 4: Check for Duplicate Records

We verify there are no duplicate NEO IDs in the dataset. Duplicates could inflate model
performance by leaking the same object into both training and test sets.

```{r duplicates}
sum(duplicated(neo_clean$neo_reference_id))
```

No duplicates found. Each row represents a unique NEO.

### Step 5: Final Cleaned Dataset

```{r final-clean}
glimpse(neo_clean)
```

---

## Visualizations

### Class Balance

```{r class-balance}
neo_clean %>%
  count(hazardous) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = hazardous, y = n, fill = hazardous)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.1)), vjust = -0.4) +
  scale_fill_manual(values = class_colors) +
  labs(
    title = "Class Distribution of Hazardous Label",
    caption = "94% of NEOs in this dataset are non-hazardous: a significant class imbalance.",
    x = NULL,
    y = "Count"
  ) +
  theme_minimal()
```

Nearly 94% of objects in the cleaned dataset are classified as non-hazardous. This class
imbalance is important: a naive model that predicts "Not Hazardous" for every object
would still achieve ~94% accuracy. Throughout the analysis we pay special attention to
recall (the proportion of actual hazardous asteroids the model correctly identifies)
rather than accuracy alone.

### Distributions by Hazard Class

```{r distributions-by-class}
plot_df <- neo_clean %>%
  select(hazardous, diam_avg_km, relative_velocity_km_s, miss_distance_km) %>%
  pivot_longer(
    cols = c(diam_avg_km, relative_velocity_km_s, miss_distance_km),
    names_to = "feature",
    values_to = "value"
  ) %>%
  mutate(
    feature = recode(
      feature,
      diam_avg_km = "Average Diameter (km)",
      relative_velocity_km_s = "Relative Velocity (km/s)",
      miss_distance_km = "Miss Distance (km)"
    )
  )

ggplot(plot_df, aes(x = value, fill = hazardous, color = hazardous)) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  facet_wrap(~ feature, scales = "free_x", ncol = 1) +
  scale_x_log10(labels = scales::label_number()) +
  scale_fill_manual(values = class_colors) +
  scale_color_manual(values = class_colors) +
  labs(
    title = "Feature Distributions by Hazard Class (Log X Scale)",
    caption = "Diameter shows the clearest separation; velocity and miss distance overlap substantially.",
    x = NULL,
    y = "Density",
    fill = "Class",
    color = "Class"
  ) +
  theme_minimal()
```

Diameter shows the clearest separation between the two classes: hazardous asteroids
cluster toward larger sizes. Relative velocity and miss distance distributions overlap
heavily, suggesting they are weaker standalone predictors.

### Relative Velocity vs. Miss Distance

```{r velocity-vs-miss}
ggplot(
  neo_clean,
  aes(x = relative_velocity_km_s, y = miss_distance_km, color = hazardous)
) +
  geom_point(alpha = 0.45, size = 1.6) +
  scale_y_log10(labels = scales::label_number()) +
  scale_color_manual(values = class_colors) +
  labs(
    title = "Relative Velocity vs. Miss Distance",
    caption = "Hazardous and non-hazardous NEOs are largely interleaved; neither variable alone separates the classes.",
    x = "Relative Velocity (km/s)",
    y = "Miss Distance (km, log scale)",
    color = "Class"
  ) +
  theme_minimal()
```

Hazardous and non-hazardous objects are largely interleaved across the velocity-distance
space, reinforcing that these two variables are not sufficient on their own to classify
hazard status. The strongest visual patterns appear when we look at size-related variables.

### Correlation Heatmap (Numeric Predictors)

```{r correlation-heatmap}
cor_df <- neo_clean %>%
  select(
    absolute_magnitude_h,
    est_diam_min_km,
    est_diam_max_km,
    diam_avg_km,
    miss_distance_km,
    relative_velocity_km_s,
    orbital_eccentricity,
    orbital_inclination,
    orbital_semi_major_axis
  )

cor_mat <- cor(cor_df, use = "complete.obs")

cor_long <- as.data.frame(as.table(cor_mat)) %>%
  rename(var1 = Var1, var2 = Var2, correlation = Freq)

ggplot(cor_long, aes(x = var1, y = var2, fill = correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Correlation Heatmap of Numeric Predictors",
    caption = "Absolute magnitude and diameter are highly negatively correlated: they measure the same underlying property.",
    x = NULL,
    y = NULL,
    fill = "r"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The heatmap reveals a near-perfect negative correlation between absolute magnitude and
both diameter columns; these variables measure the same underlying physical property
(size) from different angles. This multicollinearity means the model may attribute
importance to one or the other somewhat arbitrarily; we treat them as overlapping evidence
rather than independent signals.

### Variable Importance (Classification Tree)

```{r variable-importance}
tree_fit <- rpart(
  hazardous ~ diam_avg_km + relative_velocity_km_s + miss_distance_km +
    absolute_magnitude_h + orbital_eccentricity + orbital_inclination +
    orbital_semi_major_axis,
  data = neo_clean,
  method = "class"
)

importance_df <- tibble(
  variable = names(tree_fit$variable.importance),
  importance = as.numeric(tree_fit$variable.importance)
) %>%
  arrange(desc(importance))

ggplot(importance_df, aes(x = reorder(variable, importance), y = importance)) +
  geom_col(fill = "#1B9E77") +
  coord_flip() +
  labs(
    title = "Variable Importance from Classification Tree",
    x = NULL,
    y = "Importance"
  ) +
  theme_minimal()
```

This first tree is an exploratory check rather than the final model evaluation. A proper
held-out train/test comparison appears after the remaining research-question plots.

> **Note on exploration:** We initially explored scatter plots of raw velocity and miss
> distance before realizing the overlap between classes was too heavy to be informative
> without a log scale. We also attempted histograms for the orbital variables before
> settling on boxplots, which made the median differences between classes much clearer.
> The pairwise scatter panel went through several variable combinations; the four shown
> were selected because they produced the most visible class separation.

### Orbital Characteristics by Hazard Class

```{r orbital-boxplots}
neo_clean %>%
  select(hazardous, orbital_eccentricity, orbital_inclination, orbital_semi_major_axis) %>%
  # log10 transform semi-major axis before pivoting: raw values extend to 300+ AU,
  # which collapses the box to nothing; log scale reveals the actual distribution
  mutate(orbital_semi_major_axis = log10(orbital_semi_major_axis)) %>%
  pivot_longer(
    cols = c(orbital_eccentricity, orbital_inclination, orbital_semi_major_axis),
    names_to = "feature",
    values_to = "value"
  ) %>%
  mutate(
    feature = recode(
      feature,
      orbital_eccentricity    = "Orbital Eccentricity",
      orbital_inclination     = "Orbital Inclination (deg)",
      orbital_semi_major_axis = "Semi-Major Axis (log10 AU)"
    )
  ) %>%
  ggplot(aes(x = hazardous, y = value, fill = hazardous)) +
  geom_boxplot(outlier.alpha = 0.15, outlier.size = 0.8) +
  facet_wrap(~ feature, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = class_colors) +
  labs(
    title = "Orbital Characteristics by Hazard Classification",
    caption = "Hazardous NEOs show higher orbital eccentricity; inclination and semi-major axis differ less between classes.",
    x = NULL,
    y = NULL,
    fill = "Class"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Hazardous NEOs have a noticeably higher median orbital eccentricity than non-hazardous
ones; their orbits are more elongated, increasing the chance they cross Earth's path.
Orbital inclination and semi-major axis show less separation, suggesting eccentricity
is the more informative orbital predictor.

### Absolute Magnitude by Hazard Class

```{r magnitude-density}
ggplot(neo_clean, aes(x = absolute_magnitude_h, fill = hazardous, color = hazardous)) +
  geom_density(alpha = 0.3, linewidth = 0.9) +
  scale_fill_manual(values = class_colors) +
  scale_color_manual(values = class_colors) +
  labs(
    title = "Distribution of Absolute Magnitude by Hazard Classification",
    caption = "Hazardous NEOs concentrate at lower magnitude values (brighter, larger objects).",
    x = "Absolute Magnitude (H)",
    y = "Density",
    fill = "Class",
    color = "Class"
  ) +
  theme_minimal()
```

```{r magnitude-boxplot}
ggplot(neo_clean, aes(x = hazardous, y = absolute_magnitude_h, fill = hazardous)) +
  geom_boxplot(outlier.alpha = 0.2, outlier.size = 0.8, show.legend = FALSE) +
  scale_fill_manual(values = class_colors) +
  labs(
    title = "Absolute Magnitude by Hazard Classification",
    caption = "The median H value is several units lower for hazardous NEOs, confirming a strong size signal.",
    x = NULL,
    y = "Absolute Magnitude (H)"
  ) +
  theme_minimal()
```

Both plots confirm that hazardous NEOs cluster at lower absolute magnitude values,
meaning they tend to be brighter and physically larger. Given the near-perfect correlation
between magnitude and diameter shown in the heatmap, these two variables provide
overlapping rather than independent evidence about size.

### Pairwise Scatterplots of Top Predictors

```{r pairwise-scatter}
var_pairs <- list(
  list(x = "diam_avg_km",          y = "absolute_magnitude_h"),
  list(x = "diam_avg_km",          y = "orbital_eccentricity"),
  list(x = "diam_avg_km",          y = "orbital_inclination"),
  list(x = "absolute_magnitude_h", y = "orbital_eccentricity"),
  list(x = "absolute_magnitude_h", y = "orbital_inclination"),
  list(x = "orbital_eccentricity", y = "orbital_inclination")
)

pair_df <- bind_rows(lapply(var_pairs, function(p) {
  neo_clean %>%
    transmute(
      hazardous,
      x_val   = .data[[p$x]],
      y_val   = .data[[p$y]],
      panel   = paste(p$x, "vs", p$y)
    )
}))

ggplot(pair_df, aes(x = x_val, y = y_val, color = hazardous)) +
  geom_point(alpha = 0.2, size = 0.6) +
  facet_wrap(~ panel, scales = "free", ncol = 3) +
  scale_color_manual(values = class_colors) +
  labs(
    title = "Pairwise Scatterplots of Top Four Predictors",
    caption = "Diameter vs. magnitude shows the clearest cluster separation; no single pair produces a clean boundary.",
    x = NULL,
    y = NULL,
    color = "Class"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text      = element_text(size = 7),
    axis.text       = element_text(size = 6)
  )
```

### Hazard Rate by Asteroid Size

```{r hazard-rate-by-size}
size_hazard_summary <- neo_clean %>%
  mutate(
    size_group = cut(
      diam_avg_km,
      breaks = c(0, 0.03, 0.10, 0.30, 1.00, Inf),
      labels = c("< 0.03", "0.03-0.10", "0.10-0.30", "0.30-1.00", "1.00+"),
      right = FALSE
    )
  ) %>%
  group_by(size_group) %>%
  summarise(
    n = n(),
    hazardous_n = sum(hazardous == "Hazardous"),
    hazard_rate = hazardous_n / n,
    .groups = "drop"
  )

size_hazard_summary %>%
  mutate(hazard_rate = format_percent(hazard_rate)) %>%
  rename(
    `Average Diameter Group (km)` = size_group,
    `Asteroid Count` = n,
    `Hazardous Count` = hazardous_n,
    `Percent Hazardous` = hazard_rate
  ) %>%
  kable(caption = "Hazard rate by average diameter group")

ggplot(size_hazard_summary, aes(x = size_group, y = hazard_rate)) +
  geom_col(fill = accent_color, width = 0.7) +
  geom_text(
    aes(label = scales::percent(hazard_rate, accuracy = 0.1)),
    vjust = -0.35,
    size = 3.4
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, max(size_hazard_summary$hazard_rate) * 1.15)
  ) +
  labs(
    title = "Hazard Rate by Average Diameter Group",
    x = "Average Diameter Group (km)",
    y = "Percent Hazardous"
  ) +
  theme_minimal()
```

The two smallest diameter groups have a 0.0% hazardous rate. The hazardous rate rises
sharply once average diameter reaches 0.10-0.30 km, peaks in the 0.30-1.00 km group, and
stays high for asteroids above 1.00 km. This supports the presentation takeaway that
NASA's hazardous label is strongly connected to asteroid size.

### Diameter and Absolute Magnitude Together

```{r diameter-vs-magnitude}
magnitude_diameter_corr <- cor(
  neo_clean$absolute_magnitude_h,
  neo_clean$diam_avg_km,
  use = "complete.obs"
)

tibble(
  relationship = "Absolute magnitude H vs. average diameter",
  correlation = round(magnitude_diameter_corr, 3)
) %>%
  kable(caption = "Correlation between absolute magnitude and average diameter")

ggplot() +
  geom_point(
    data = filter(neo_clean, hazardous == "Not Hazardous"),
    aes(x = absolute_magnitude_h, y = diam_avg_km, color = hazardous),
    alpha = 0.12,
    size = 0.8
  ) +
  geom_point(
    data = filter(neo_clean, hazardous == "Hazardous"),
    aes(x = absolute_magnitude_h, y = diam_avg_km, color = hazardous),
    alpha = 0.7,
    size = 1.1
  ) +
  scale_y_log10(labels = scales::label_number()) +
  scale_color_manual(values = class_colors) +
  labs(
    title = "Average Diameter vs. Absolute Magnitude by Hazard Class",
    x = "Absolute Magnitude H",
    y = "Average Diameter (km, log scale)",
    color = "Class"
  ) +
  theme_minimal()
```

Average diameter and absolute magnitude move together very strongly: lower absolute
magnitude generally means a brighter and larger asteroid. That is why both variables
show up as important in the analysis. For interpretation, they should be treated as
overlapping evidence about size rather than as two independent findings.

## Modeling

We use a classification tree because the earlier plots suggest threshold-like behavior:
very small asteroids are rarely labeled hazardous, while larger objects are much more
likely to be flagged. To keep the evaluation honest, the data is split into stratified
training and test sets before fitting the models.

There is an important interpretation limit: NASA's potentially hazardous asteroid label
is partly tied to object size and brightness. Since estimated diameter and absolute
magnitude are also closely related to each other, a model using those fields is partly
recovering the structure of the label rather than discovering a fully independent warning
rule. The full model is therefore best read as an explanatory model for this dataset, not
as an independent asteroid warning system.

```{r train-test-split}
model_vars_full <- c(
  "absolute_magnitude_h",
  "diam_avg_km",
  "miss_distance_km",
  "relative_velocity_km_s",
  "orbital_eccentricity",
  "orbital_inclination",
  "orbital_semi_major_axis"
)

set.seed(2026)

neo_model <- neo_clean %>%
  select(neo_reference_id, hazardous, all_of(model_vars_full)) %>%
  mutate(row_id = row_number())

train_ids <- neo_model %>%
  group_by(hazardous) %>%
  slice_sample(prop = 0.80) %>%
  ungroup() %>%
  pull(row_id)

train_data <- neo_model %>%
  filter(row_id %in% train_ids) %>%
  select(-row_id, -neo_reference_id)

test_data <- neo_model %>%
  filter(!row_id %in% train_ids) %>%
  select(-row_id, -neo_reference_id)

tibble(
  split = c("Train", "Test"),
  rows = c(nrow(train_data), nrow(test_data)),
  hazardous_rate = c(
    mean(train_data$hazardous == "Hazardous"),
    mean(test_data$hazardous == "Hazardous")
  )
) %>%
  mutate(hazardous_rate = format_percent(hazardous_rate)) %>%
  kable(caption = "Stratified train/test split")
```

### Evaluation Functions

```{r metric-functions}
safe_divide <- function(numerator, denominator) {
  ifelse(denominator == 0, NA_real_, numerator / denominator)
}

confusion_metrics <- function(actual, predicted, positive = "Hazardous") {
  actual <- as.character(actual)
  predicted <- as.character(predicted)
  negative <- "Not Hazardous"

  tp <- sum(actual == positive & predicted == positive)
  tn <- sum(actual == negative & predicted == negative)
  fp <- sum(actual == negative & predicted == positive)
  fn <- sum(actual == positive & predicted == negative)

  tibble(
    tp = tp,
    tn = tn,
    fp = fp,
    fn = fn,
    accuracy = safe_divide(tp + tn, tp + tn + fp + fn),
    recall = safe_divide(tp, tp + fn),
    specificity = safe_divide(tn, tn + fp),
    precision = safe_divide(tp, tp + fp)
  )
}
```

### Fit Classification Tree and Logistic Regression

To verify that the classification tree findings are not an artifact of the model type,
we fit a logistic regression on the same training data and compare performance on the
same test set. This follows the principle that unintuitive or convenient-looking results
should be checked with at least one alternative approach before being reported as findings.

```{r fit-models}
full_formula <- as.formula(
  paste("hazardous ~", paste(model_vars_full, collapse = " + "))
)

tree_full <- rpart(
  full_formula,
  data = train_data,
  method = "class",
  control = rpart.control(cp = 0.001, minbucket = 25)
)

tree_full_predicted <- predict(tree_full, newdata = test_data, type = "class")

# Logistic regression on the same split for comparison
logit_data_train <- train_data %>%
  mutate(hazardous_bin = as.integer(hazardous == "Hazardous"))

logit_data_test <- test_data %>%
  mutate(hazardous_bin = as.integer(hazardous == "Hazardous"))

logit_formula <- as.formula(
  paste("hazardous_bin ~", paste(model_vars_full, collapse = " + "))
)

logit_fit <- glm(logit_formula, data = logit_data_train, family = binomial)

logit_prob <- predict(logit_fit, newdata = logit_data_test, type = "response")
logit_predicted <- factor(
  ifelse(logit_prob >= 0.5, "Hazardous", "Not Hazardous"),
  levels = c("Not Hazardous", "Hazardous")
)
```

## Results

The classification tree and logistic regression were evaluated on the same held-out test
set using four metrics. **Accuracy** is the overall proportion of correct predictions,
but is inflated here by the class imbalance since always predicting "Not Hazardous" would
score around 94%. **Recall** (also called sensitivity or true positive rate) is the
proportion of actual hazardous asteroids that the model correctly flags; this is the most
important metric for a planetary defense context where missing a real threat is costly.
**Specificity** measures how well the model avoids false alarms on non-hazardous objects.
**Precision** measures how often the model is right when it does predict hazardous.

### Classification Tree Performance

```{r model-comparison}
model_metrics <- confusion_metrics(test_data$hazardous, tree_full_predicted) %>%
  select(
    accuracy,
    recall,
    specificity,
    precision,
    tp,
    tn,
    fp,
    fn
  )

model_metrics %>%
  mutate(
    across(c(accuracy, recall, specificity, precision), ~ format_percent(.x))
  ) %>%
  select(
    Accuracy = accuracy,
    Recall = recall,
    Specificity = specificity,
    Precision = precision,
    `True Positives` = tp,
    `True Negatives` = tn,
    `False Positives` = fp,
    `False Negatives` = fn
  ) %>%
  kable(caption = "Classification tree performance on the test data")
```

The full tree has high specificity, meaning it is good at recognizing non-hazardous
asteroids, but recall is the key limitation. Because the hazardous class is small,
accuracy can look strong even when many hazardous asteroids are missed. This is why the
confusion matrix below is important: it shows how many hazardous asteroids the model
catches and misses.

### Confusion Matrix Heatmap

```{r confusion-heatmap}
confusion_df <- tibble(
  actual = test_data$hazardous,
  predicted = tree_full_predicted
) %>%
  count(actual, predicted) %>%
  complete(
    actual = levels(test_data$hazardous),
    predicted = levels(test_data$hazardous),
    fill = list(n = 0)
  )

ggplot(confusion_df, aes(x = predicted, y = actual, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), color = "white", fontface = "bold") +
  scale_fill_gradient(low = "#C6DBEF", high = class_colors[["Not Hazardous"]]) +
  labs(
    title = "Confusion Matrix: Classification Tree",
    x = "Predicted Class",
    y = "Actual Class",
    fill = "Count"
  ) +
  theme_minimal()
```

### Model Comparison: Tree vs. Logistic Regression

```{r logit-comparison}
tree_metrics <- confusion_metrics(test_data$hazardous, tree_full_predicted) %>%
  mutate(model = "Classification Tree")

logit_metrics <- confusion_metrics(test_data$hazardous, logit_predicted) %>%
  mutate(model = "Logistic Regression")

bind_rows(tree_metrics, logit_metrics) %>%
  mutate(across(c(accuracy, recall, specificity, precision), ~ format_percent(.x))) %>%
  select(
    Model = model,
    Accuracy = accuracy,
    Recall = recall,
    Specificity = specificity,
    Precision = precision,
    TP = tp,
    TN = tn,
    FP = fp,
    FN = fn
  ) %>%
  kable(caption = "Classification tree vs. logistic regression on the same test set")
```

Both models show similar patterns: high specificity, moderate-to-low recall, and
accuracy figures that are partly inflated by the class imbalance. The close agreement
between the two very different model types strengthens confidence that the findings
reflect genuine structure in the data rather than a quirk of the decision tree algorithm.
Where the models disagree, the difference is small and consistent with sampling variation
on the minority class.

### Classification Tree Variable Importance

```{r model-variable-importance}
importance_df <- tibble(
  variable = names(tree_full$variable.importance),
  importance = as.numeric(tree_full$variable.importance)
) %>%
  mutate(variable = recode(variable, !!!feature_labels)) %>%
  arrange(desc(importance))

importance_df %>%
  mutate(importance = round(importance, 1)) %>%
  kable(caption = "Most important variables in the full classification tree")

ggplot(importance_df, aes(x = reorder(variable, importance), y = importance)) +
  geom_col(fill = accent_color) +
  coord_flip() +
  labs(
    title = "Variable Importance from Full Classification Tree",
    x = NULL,
    y = "Importance"
  ) +
  theme_minimal()
```

The model results reinforce the visual findings: size and brightness are central to the
hazardous label, while the other orbital and close-approach variables provide weaker
separation. The full tree is useful as an explanatory check, but its low recall means it
should not be interpreted as a real-world asteroid warning system.

> **Skepticism note:** The strong performance on specificity and the weaker recall are
> consistent across both the classification tree and logistic regression. This consistency
> across two fundamentally different algorithms reduces the chance that the pattern is a
> model-specific artifact. However, both models share the same predictor set and the same
> label definition issue described above, so the agreement does not resolve the underlying
> concern about circular prediction. A more rigorous test would require a dataset where
> the hazardous label was assigned independently of the size and brightness fields used
> as predictors.

## Conclusions

Our analysis of 42,919 Near-Earth Objects leads to three main findings.

**Size and brightness are the dominant predictors.** Hazardous asteroids are consistently
larger and brighter than non-hazardous ones, and the hazard rate jumps sharply once
average diameter exceeds 0.10 km. Absolute magnitude and average diameter are nearly
redundant predictors; both are proxies for physical size, and either alone carries most
of the predictive signal.

**Orbital eccentricity adds a secondary signal.** Hazardous NEOs have more elliptical
orbits on average, which makes physical sense: highly eccentric orbits are more likely to
cross Earth's path. Inclination and semi-major axis contribute less and were less
important in the fitted tree.

**The class imbalance limits practical recall.** The full classification tree achieves
high specificity (it is good at ruling out non-hazardous objects) but struggles to
catch all hazardous asteroids because they represent only ~6% of the dataset. Accuracy alone is misleading in this setting; recall is the more important metric for a real
planetary defense application.

### Limitations

The model should be interpreted as an explanatory tool for this dataset, not as a
real-world asteroid warning system. NASA's Potentially Hazardous Asteroid definition
is partly based on object size and orbital distance from Earth, the same properties
we use as predictors, so the model is partly recovering the structure of the label
rather than discovering a fully independent rule.

### Future Work

Several directions would strengthen this analysis:

- **Add Earth MOID** (Minimum Orbital Intersection Distance), the orbital metric that
  is formally part of the PHA definition and is not currently in the dataset
- **Address class imbalance** through oversampling (SMOTE) or cost-sensitive learning
  to improve recall on hazardous objects
- **Compare model types**: logistic regression and random forests would provide useful
  benchmarks against the classification tree
- **Track the same asteroid over multiple close approaches** rather than treating each
  approach record independently, which could introduce subtle data leakage
- **Drop one of diameter or absolute magnitude** and test whether predictive performance
  degrades, to determine whether both variables are truly needed
